### 2011-02-09 paper/abstract.tex
First ugly version of report in acm style.  Figures ugly, bibliography commented out (not yet working)
Restructuring of report into multiple files to minimize conflicts, references working, dealt old sections to appropriate locations in paper framework.
lots of editing of the intro/motivation/previouswork, and some in the robot platform

This paper presents an array of approaches to optimizing a quadrupedal
gait for forward speed.  We implement, test, and compare different
learning strategies including uniform and Gaussian random hill
climbing, policy gradient reinforcement learning\cite{kohl},
Nelder-Mead simplex\cite{nm}, new predictive methods based on linear
and support vector regression, and an evolved neural network
(HyperNEAT)\cite{clune}.  We compare results to a baseline random
search method.  Many of the methods resulted in walks significantly
faster than previously hand-tuned gaits.

@@

Keywords: Learning Control, Walking Robots, Gait Learning, Quadruped

>>>>

This paper presents an array of approaches to optimizing a quadrupedal
gait for forward speed.  We implement, test, and compare different
learning strategies including uniform and Gaussian random hill
climbing, policy gradient reinforcement learning,
Nelder-Mead simplex, new predictive methods based on linear
and support vector regression, and an evolved neural network
(HyperNEAT).  We compare results to a baseline random
search method.  Many of the methods resulted in walks significantly
faster than previously hand-tuned gaits.

@@

\edit{Write the abstract}

### 2011-02-09 paper/acknowledgments.tex
lots of editing of the intro/motivation/previouswork, and some in the robot platform

Thanks to ..., some grant maybe, Jim T\o rreson, ...and the reviewers.

\edit{write this, old below}

% List any people not on the team who helped you with your project: Your
% TA, other people you consulted with or had useful discussions (say in 
% a word or two what they did), people who proofread your report, and 
% any external code you used (libraries etc).
\begin{itemize}
\item Hod Lipson, Cornell Computational Synthesis Lab: adviser.
\item Jim T\o rreson, University of Oslo: adviser.
\item Juan Zagal, University of Chile: designed and printed robot and provided code for hand tuned gait.
\item Jeff Clune, CCSL Red Couch: collaborated on HyperNEAT implementation/testing.
\item Cooper Bills, Cornell University: assisted with Wii tracker development.
\item Anshumali Srivastava, Cornell University: Teaching Assistant.
\end{itemize}

>>>>

Left blank for anonymity.

% List any people not on the team who helped you with your project: Your
% TA, other people you consulted with or had useful discussions (say in 
% a word or two what they did), people who proofread your report, and 
% any external code you used (libraries etc).
%\begin{itemize}
%\item Hod Lipson, Cornell Computational Synthesis Lab: adviser.
%\item Jim T\o rreson, University of Oslo: adviser.
%\item Juan Zagal, University of Chile: designed and printed robot and provided code for hand tuned gait.
%\item Jeff Clune, CCSL Red Couch: collaborated on HyperNEAT implementation/testing.
%\item Cooper Bills, Cornell University: assisted with Wii tracker development.
%\item Anshumali Srivastava, Cornell University: Teaching Assistant.
%\end{itemize}

### 2011-02-09 paper/introduction.tex
lots of editing of the intro/motivation/previouswork, and some in the robot platform

\acmFig{robot_close.jpg}{1}{The 3D printed quadruped robot used for
  this study.}

Gaits for walking robots are often designed with some explicit or
implicit goal in mind.  For some applications, the design criteria may
be obvious --- perhaps the robot needs to move as quickly or as
efficiently as possible --- but other times the objective is more complicated,
requiring simultaneous optimization of several desired traits, each
with its own relative importance.  The different combinations of desired traits and the relative weight placed on each can produce drastically different gaits.    For example, Honda's Asimo
\cite{_asmio} and Boston Dynamic's Big Dog \cite{_bigdog} both require
gaits that are relatively quick, power efficient, and robust to
changing terrain, but they vary widely in the importance placed
on each attribute.  Big Dog's gait is optimzed for much more difficult terrain than Asimo's, resulting in a gait of a completely different form.

Whatever the design goals, gaits may be obtained by one of several methods.  Gaits may be designed by hand by an expert, as in \cite{_asmio, _bigdog}, or they may be learned by repeated trial and evaluation.  Learned gaits offer several advantages over manually designed gaits.  Automatically learning gaits can save valuable engineering time and can allow customization of gaits to a particular robot and its unique actuators.  Most importantly, learned gaits can outperform manually designed gaits

save programmer time
customize gaits per model
design better gaits than people

In many cases additional objectives may be

, or with minimal power usage or strain on
actuators.  In many cases, however,

In other cases, the objective may be more complicated; perhaps it is desired that the gait be fairly [robust]

\_asimo and \_bigDog very different (and some hexapod?)

Many times gait will be designed and tweaked by hand \cite{_asimo}

here we describe a study, a comparison of few methods. started with
sped, but then not-breaking as well.

When designing gaits for walking robots, often it is desirable that
this gait enable the robot to walk as

Old: design by hand
Newer: Learn gait
  - parmametrized or more free form (HN, ???)
  - in sim or hardware
Us: first time HN in hardware, first time compared to other methods aiming for symmetry on same hardware

Often
these optimizations are done manually by an expert who designs and
tweaks a gait specifically for a given objective.  Other groups have
used learning methods to generate gaits optimized for some metric.
Approaches differ in their starting assumptions, some essentially
tweaking the parameters of a hand-tuned model \cite{chernova}, others
exploring a reasonably compact parameter space \cite{kohl}, and still
others beginning with few assumptions besides periodicity
\cite{zykov}.

\edit{Write this}

%\editbox{Outline \\
% - Why gait learning is cool \\
% - Gait learning has been done before, in simulation and on hardware (cite, cite, cite) \\
% - HyperNEAT has been used to evolve gaits before in simulation~{clune2009evolving, clune2011performance, clune2009hybrid}, and it worked well in part because the gaits were regular, raising the question of whether this high performance was just because the gaits were regular? We test this hypothesis by comparing HyperNEAT to traditional machine learning methods biased towards regular gaits.  \\
% - Here, we do two cool things for the first time: \\
% - - evolve HN in hardware \\
% - - compare it to other learning methods on the same hardware.}

\edit{Old Quadratot intro follows}

% Motivate and abstractly describe the problem you are addressing and
% how you are addressing it. What is the problem? Why is it important?
% What is your basic approach? A short discussion of how it fits into
% % related work in the area is also desirable. Summarize the basic
% results and conclusions that you will present.

Applications of walking robots often call for the ability to walk as
quickly, efficiently, or with as little power as possible.  Often
these optimizations are done manually by an expert who designs and
tweaks a gait specifically for a given objective.  Other groups have
used learning methods to generate gaits optimized for some metric.
Approaches differ in their starting assumptions, some essentially
tweaking the parameters of a hand-tuned model \cite{chernova}, others
exploring a reasonably compact parameter space \cite{kohl}, and still
others beginning with few assumptions besides periodicity
\cite{zykov}.

We aimed to strike a middle ground between these approaches.  Our
motion generator did not rely on hand-tweaked gaits, but it did use
parameterized gaits which, by their nature, impose some assumptions on
the answers produced.  We then used machine learning to design gaits
for a quadruped robot with these models.  This paper presents a
comparison of the different learning methods implemented.  Most
methods created walks that are several times faster than the original
hand-tuned gait.  We invite readers with short attention spans to view
a video of some of our results online here:

%\url{http://www.youtube.com/watch?v=ODoiOj9DdGg}
\texttt{http://www.youtube.com/watch?v=ODoiOj9DdGg}

%The code so far consists of the robot class, optimization class,
%parameterized model, sine model, and camera feedback.

>>>>

\acmFig{robot_close.jpg}{1}{The quadruped robot. The translucent parts were created with 3D-printing technology.}

Gaits for walking robots are often designed with some explicit or
implicit goal in mind.  For some applications, the design criteria may
be obvious --- perhaps the robot needs to move as quickly or as
efficiently as possible --- but other times the objective is more complicated,
requiring simultaneous optimization of several desired traits, each
with its own relative importance.  The different combinations of desired traits and the relative weight placed on each can produce drastically different gaits.    For example, Honda's Asimo
\cite{chestnutt2006footstep} and Boston Dynamic's Big Dog \cite{raibert2008bigdog} both require
gaits that are relatively quick, power efficient, and robust to
changing terrain, but they vary widely in the importance placed
on each attribute.  Big Dog's gait is optimized for much more difficult terrain than Asimo's, resulting in a gait of a completely different form.

Once the design goals are decided upon, gaits may be obtained by one of several methods.  Gaits may be designed manually by an expert, as for Asimo and Big Dog \cite{chestnutt2006footstep, raibert2008bigdog}.  Alternately, they may be learned through repeated trial and evaluation.  Learned gaits offer several advantages over manually designed gaits.  Automatically learning gaits can save valuable engineering time and can allow customization of gaits to a particular robot and its unique actuators.  Most importantly, in some cases learned gaits can outperform manually designed gaits~\cite{hornby2005autonomous, valsalam2008modular}.

In this paper we compare the performance of three different methods of designing gaits: manual design, parametrized gaits optimized with seven different learning methods, and gaits generated by evolving neural networks with the HyperNEAT generative encoding~\cite{stanley2009hypercube}. While some of these methods, such as HyperNEAT, have been tested in simulation~\cite{clune2009evolving, clune2011performance}, we investigate how they perform when evolving on a physical robot (\figref{robot_close.jpg}).

Previous work has shown that quadruped gaits perform better when they are \emph{regular} (i.e.\ when the legs are coordinated)~\cite{clune2009evolving, clune2011performance,valsalam2008modular}. For example, HyperNEAT produced fast, natural gaits in part because its bias towards regular gaits created coordinated movements that outperformed gaits evolved by an encoding not biased towards regularity~\cite{clune2009evolving, clune2011performance}. One of the motivations of this paper is to investigate whether any learning method biased towards regularity would perform well at producing quadruped gaits, or whether HyperNEAT's high performance is due to additional factors, such as its abstraction of biological development (described below). We test this hypothesis by comparing HyperNEAT to seven machine learning algorithms biased toward regularity.

A final motivation of this problem is to simply evolve effective gaits for a physical robot. Because HyperNEAT performed well in simulation, it is interesting to test whether it can produce a fast gait for a physical robot. It is additionally interesting to test how more traditional machine learning techniques compete with evolutionary algorithms when evolving in hardware.

### 2011-02-09 paper/platformDetails.tex
lots of editing of the intro/motivation/previouswork, and some in the robot platform

Computation for gait learning, fitness evaluation, and robot control
is performed on the compact on-board CompuLab Fit-PC2, running Ubuntu
Linux.  All gait generation, learning, and fitness evaluation code,
except HyperNEAT, is written in Python and is available on our website
\cite{quadraWeb}.  HyperNEAT is written in C++.  We made use of the
excellent \code{pydynamixel} library \cite{pydynamixel} for
controlling the servos, available under the GPLv2 license.  The robot
was set up to connect to a wireless network on boot, and thus we were
able to control all experiments via ssh.

@@

To track the position of the robot and determine gait fitness, we
mounted a Nintendo Wii remote on the ceiling and an infrared LED on
top of the robot (\figref{wiiMote.JPG}).  The Wii remote contains an
IR camera that can track and report the position of any IR sources in
its image frame.  The resolution of the camera was 1024 by 768 pixels,
which produced a resolution of 1.7mm per pixel when mounted at a
height of \edit{???}m.

A separate tracking server, written in Python, ran on the robot and
interfaced with the Wii remote via bluetooth using the CWiid
library\cite{cwiid}.  Our fitness testing code connected to this
server via socket and requested position updates at the beginning and
end of each run (see \secref{fitnessEvaluation for more details}).

@@

To run a gait on the robot, one simply provides a \emph{gait
  function}, any Python function that accepts a single input --- time
starting at 0 --- and outputs a list of nine commanded positions, one
for each servo.  To safeguard against limb collision with the robot
body, the robot control code cropped the commands to a safe range.
This range was [150, 770] for the inner leg servos, [30, 680] for the
outer leg servos, and [392, 623] for the center hip servo.

>>>>

All of the computation for gait learning, fitness evaluation, and robot control
was performed on the compact on-board CompuLab Fit-PC2, running Ubuntu
Linux.  All gait generation, learning, and fitness evaluation code,
except HyperNEAT, is in Python and is available on our website
\cite{quadraWeb}.  HyperNEAT is written in C++.  We controlled the servos with the \code{pydynamixel} library \cite{pydynamixel}.  The robot
connects to a wireless network on boot, which enabled us to control it via SSH.

@@

To track the position of the robot and thus determine gait fitness, we
mounted a Nintendo Wii remote on the ceiling and an infrared LED on
top of the robot (\figref{wiiMote_crop.jpg}).  The Wii remote contains an
IR camera that can track and report the position of any IR sources in
its image frame.  The resolution of the camera was 1024 by 768 pixels,
which produced a resolution of 1.7mm per pixel when mounted at a
height of \edit{???}m.

A separate tracking server, written in Python, ran on the robot and
interfaced with the Wii remote via bluetooth using the CWiid
library\cite{cwiid}.  Our fitness testing code connected to this
server via socket and requested position updates at the beginning and
end of each run (see \secref{fitnessEvaluation}).

@@

To run a gait on the robot, a \emph{gait
  function} is provided, any Python function that accepts a single input --- time
starting at 0 --- and outputs a list of nine commanded positions, one
for each servo.  To safeguard against limb collision with the robot
body, the robot control code cropped the commands to a safe range.
This range was [150, 770] for the inner leg servos, [30, 680] for the
outer leg servos, and [392, 623] for the center hip servo.

### 2011-02-09 paper/relatedWork.tex
lots of editing of the intro/motivation/previouswork, and some in the robot platform

\editbox{If we talk about past gait-learning work in here, is it a bit redundant?}

\edit{write this.  Old quadratot section below}

% Briefly explain who else worked on related problems in the past and what
% methods they used. Explain if you are using similar methods, or if your
% approach is different and if so - how (either is ok).

Various maching learning techniques have proven to be useful in
finding control policies for a wide variety of robots. Kohl and
Stone\cite{kohl} presented a policy gradient reinforcement learning
approach for generating a fast walk on legged robots. We experimented
with this method to create a walk for our robot (Policy Gradient
Descent). Chernova and Velosa\cite{chernova} took an evolutionary
approach to this problem which we did not implement.  Zykov, Bongard,
and Lipson\cite{zykov} describe the evolution of dynamic gaits on a
physical robot requiring no prior assumptions about the locomotion
pattern beyond the fact that it should be rhythmic.

>>>>

Various machine learning techniques have proven to be effective at generating gaits for legged robots. Kohl and Stone presented a policy gradient reinforcement learning approach for generating a fast walk on legged robots\cite{kohl}. We experiment with this method to create a walk for our robot (called Policy Gradient Descent, described below). Others have evolved gaits for legged robots, producing competitive results~\cite{chernova2005evolutionary, hornby2005autonomous, zykov, clune2009evolving, clune2011performance, clune2009hybrid, clune2009sensitivity, tellez2006evolving, valsalam2008modular}. In fact, an evolved gait was used in the first commercially-available version of Sony's AIBO robot~\cite{hornby2005autonomous}. Except for work with HyperNEAT~\cite{clune2009evolving, clune2011performance, clune2009hybrid, clune2009sensitivity}, the previous evolutionary approaches have helped evolution 'see' the regularity of the problem by manually decomposing the task. In other words, experimenters have to choose which legs should be coordinated, or otherwise facilitate the coordination of robotic legs. Part of the motivation of this paper is to see if the space of regularities can be explored automatically, which has previously performed well~\cite{valsalam2008modular}, and to perform a direct comparison on the same robot to HyperNEAT.
